Master's Project Journal

### 2019-04-07 thesis/Background.tex
sunday funday pt1

%DOWN TO HERE DONE FOR ACRO
\section{Multi-oscillator Designs}
The designs described previously, are all similar in that they have a single central oscillator that provides the clock for all areas of the chip, whereas the following methods attempt to synchronise multiple oscillators, each of which provides the clock for a single clocking area. The main advantage of a multi-oscillator design, is that as each clocking area has its clock created locally, there is no problem with signal degradation. Similar to a decentralised skew compensation scheme comparisons are only made to neighbouring blocks and as such there is reduced power loss due to transmission.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.7]{../mizuno1998noise}
	\caption{Coupled Oscillator Clock Delivery Circuit \cite{mizuno1998noise}.}
	\label{fig:mizuno1998noise}
\end{figure}
One such method is a network of oscillators as in Figure \ref{fig:mizuno1998noise} which uses coupled PLLs to generate local clocks. The advantage of this method is that only the control voltage $v_c$ needs delivery to all areas of the chip, thus alleviating the need for a power hungry distribution circuit, while also being more noise-immune than the transmission of a high frequency clock. However this design still suffers from clock variation as all VCOs are fed the same control voltage, and this is acknowledged by the authors:
\begin{quotation}
	\textit{Unfortunately, as with the conventional ... method, distributing the VCOs over the entire chip causes the problem that jitter and skew are increased by variations in the fabrication process (static), temperature, and power supply (dynamic)} \cite{mizuno1998noise}.
\end{quotation}
This design also only compares the phase of a single branch of the distribution network with the reference, in order calculate the control voltage which again may contribute to a degradation in clock synchronisation. %TODO why

Another potential multi-oscillator design synchronises the oscillators using the phase relationship with the oscillators in the neighbouring clocking area. Once again, this negates the requirement for a global distribution structure and comparisons need only be made with neighbouring areas. As a phase lock loop is being used only a divided down version of the clock is required. This in turn means the hardware transporting the divided clock signal to the phase comparator, has significantly lower requirements placed on it, thus lowering the power consumption. Pratt and Nguyen initially proposed this design in their paper published in 1995 entitled ``\textit{Distributed Synchronous Clocking}'' in which they propose a Cartesian grid of clocking areas, each with their own PLL, which has become known as a PLL Network \cite{pratt1995distributed}. In this design any given node is synchronised with its neighbours and one of the corner nodes is additionally synchronised with the reference. According to the authors this is ``a simple, effective way to achieve low cost, high quality, low skew clock generation in a synchronous parallel processor''. This methodology was implemented by Gutnik \textit{et al} (2000) who found:
\begin{quote}
	\textit{Design and measurements on this chip confirm that generating and synchronizing multiple clocks on chip is feasible. Neither the power nor the area overhead of multiple PLLs is substantial compared to the cost of distributing the clock by conventional means} \cite{gutnik2000active}.
\end{quote}
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.7]{../gutnik2000active}
	\caption{PLL Network Topology \cite{gutnik2000active}.}
	\label{fig:gutnik2000active}
\end{figure}

\section{ADPLL Networks}
As a PLL network is an analog circuit, its integration in a modern IC is a barrier to usage and it has not been used in any commercial designs \cite{zianbetov2013distributed}. An alternative design that is more suitable for current fabrication techniques eschews from using analog components and instead implements the network of controlled oscillators using only digital circuitry, hence the name All-Digital PLL. A 4x4 ADPLL network was designed and prototyped in 65 nm CMOS by Zianbetov and Shan in order to test the suitability of the technique as a clock distributor \cite{zianbetov2013phd,shan2014phd}. %TODO fix phd references

In this design the oscillators are coupled using digital phase comparators which attempt to measure the phase error between two oscillators with each non edge node being connected to four neighbours. Figure \ref{fig:eldar_node} shows high level detail of the architecture of both the entire clocking system and that of an individual node in the design. A digital PLL network has the additional advantages over its analog counterparts that it can benefit from advancements in digital circuit design suites, be reconfigurable and programmable and has a significantly greater immunity to perturbations inherent to its digital nature \cite{zianbetov2013phd}. This last advantage is of particular use in a digital environment, as otherwise there is potential for clock degradation resulting from switching of transistors.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.4]{../ccirc_2013_arch}
	\caption{Architecture of the ADPLL network and of a single node \cite{zianbetov2013distributed}.}
	\label{fig:eldar_node}
\end{figure}
Looking at the design of a given node it is notable that the function carried out by the Error Combiner is akin to a average, therefore as mentioned by Pratt and Nyugen, there is potential to settle in a stable ``mode'' in which the oscillators are not synchronised. In their paper, they present a method where initial start-up is performed uni-directionally and, once all nodes are close to alignment, full connectivity can be restored, however this was not viable in an analog system \cite{pratt1995distributed}. In creating an entirely digital system, Zianbetov and Shan could exploit the reconfigurability and and implement this system at start-up.

\section{ADPLL Architecture}
As indicated in Figure \ref{fig:mulkeen_pll} the three main building blocks of a conventional Phase Lock Loop are the Phase Detector, Loop Filter and Controllable Oscillator. In the All-Digital counterpart each of these traditionally analog components are replaced by their digital counterparts, necessitating quantisation in order to remain physically realisable. 
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.5]{../tex_files/mulkeen_pll}
	\caption{Block Diagram of a Phase Lock Loop, \textit{Wireless Systems Notes}, B. Mulkeen (2017).}
	\label{fig:mulkeen_pll}
\end{figure}

>>>>

%DOWN TO HERE DONE FOR ACRO
\section{Multi-oscillator Designs}
The designs described previously, are all similar in that they have a single central oscillator that provides the clock for all areas of the chip, whereas the following methods attempt to synchronise multiple oscillators, each of which provides the clock for a single clocking area. The main advantages of a multi-oscillator design, are that as each clocking area has its clock created locally, there is no degradation in the quality of the signal as it is distributed around the chip and the number of potential noise sources is reduced. In order to obtain global synchronisation some method of comparison between local clocking areas is required, and how this is done depends on the architecture. Regardless of the comparison is made, it is carried out between neighbouring clocking areas and as such the feedback network need not have a large footprint or power overhead.

One such method is a network of oscillators as in Figure \ref{fig:mizuno1998noise} which uses coupled \ac{PLL} to generate local clocks. Here the output of a leaf node is compared with an external reference and the operating frequency of each \ac{VCO} tuned based on the result.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.7]{../mizuno1998noise}
	\caption{Coupled Oscillator Clock Delivery Circuit \cite{mizuno1998noise}.}
	\label{fig:mizuno1998noise}
\end{figure}
The advantage of this method is the simplicity of the feedback network, requiring just the divided clock output from a single leaf node. The \ac{VCO}s then adjusted by the control voltage, $v_c$, which needs delivery to all areas of the chip. However, this is a regular signal and as such does not suffer from skew or jitter. This alleviates the need for a power hungry distribution circuit, while also being more noise-immune than the transmission of a high frequency clock. However this design still suffers from clock variation as all \ac{VCO}s are fed the same control voltage, and thus the manufacturing tolerance issues present in conventional designs persists here also. This is acknowledged by the authors:
\begin{quotation}
	\textit{Unfortunately, as with the conventional ... method, distributing the \ac{VCO}s over the entire chip causes the problem that jitter and skew are increased by variations in the fabrication process (static), temperature, and power supply (dynamic)} \cite{mizuno1998noise}.
\end{quotation}
This type of multi-oscillator design is implemented by analogue circuits, and as a result not only are the clock signals, but also the control signals are liable to variation due to noise, fabrication mismatch and power supply dynamics.

Another potential multi-oscillator clock distribution system uses the phase relationship between the oscillators driving neighbouring clock areas in order to obtain synchronisation. Once again, this negates the requirement for a global distribution structure and the signals used for comparisons need only be sent between neighbouring clocking areas. As a \ac{PLL} is being used it is again possible to perform the phase comparisons using a divided version of the generated clock. This in turn means the hardware transporting the divided clock signal to the phase comparator, has significantly lower requirements placed on it, thus lowering the power consumption due to electrical losses. Pratt and Nguyen initially proposed method of clock distribution in their 1995 paper entitled ``\textit{Distributed Synchronous Clocking}'' in which they propose a Cartesian grid of clocking areas, each with their own \ac{PLL}, which has become known as a \ac{PLL} Network \cite{pratt1995distributed}. In this design any given node is synchronised with its neighbours and one of the corner nodes is additionally synchronised with the reference. According to the authors this is ``a simple, effective way to achieve low cost, high quality, low skew clock generation in a synchronous parallel processor''. They did, however, note the presence of a phenomenon called ``mode locking'', which is setting of the network into a stable equilibrium where there are non-zero relative phases.

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.7]{../gutnik2000active}
	\caption{PLL Network Topology \cite{gutnik2000active}.}
	\label{fig:gutnik2000active}
\end{figure}

This architecture of clock distribution network was then implemented by Gutnik \textit{et al} (2000) who fabricated a 4x4 array of oscillators, operating at a centre frequency of 1.2 MHz. The oscillator was implemented as a voltage controlled ``nMOS-loaded differential ring oscillator'', and in order to mode locking the phase detector was implemented as a highly non-linear circuit. The design was a success and the authors concluded:
\begin{quote}
	\textit{Design and measurements on this chip confirm that generating and synchronizing multiple clocks on chip is feasible. Neither the power nor the area overhead of multiple \acs{PLL}s is substantial compared to the cost of distributing the clock by conventional means} \cite{gutnik2000active}.
\end{quote}
\section{ADPLL Networks}
As a \ac{PLL} network is an analog circuit, its integration in a modern \ac{IC} is a barrier to usage, and as such it has not been used in any commercial designs \cite{zianbetov2013distributed}. An alternative design that is more suitable for current fabrication techniques eschews from using analogue components and instead implements the network of controlled oscillators using only digital circuitry, hence the name All-Digital \ac{PLL}. A 4x4 \ac{ADPLL} network was designed and prototyped in 65 nm CMOS by Zianbetov and Shan in order to test the suitability of the technique as a clock distributor \cite{zianbetov2013phd,shan2014phd}.

In this design the oscillators are once again laid out in a Cartesian grid, with each node coupled to their neighbours in phase. As this is now a digital system the coupling is carried out using digital phase comparators, which attempt to measure the phase difference between two oscillators. Figure \ref{fig:eldar_node} shows high level detail of the architecture of both the entire clocking system and that of an individual node in the design. The digital nature of this architecture brings with it a number of advantages over traditional analogue implementations, as it can benefit from advancements in digital circuit design suites, be reconfigurable and programmable and has a significantly greater immunity to perturbations inherent to its digital nature, as the exact voltage of signals is of no importance \cite{zianbetov2013phd}. This last advantage is of particular use in a digital environment, as otherwise there is potential for clock degradation resulting from switching of transistors. The drawback of the switch to a digital architecture however is the presence of quantisation. Analogue designs both deliver continuous control signals to the oscillators and have a continuous phase detection capability, unlike a digital system where these actions are carried out with fixed resolution.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.4]{../ccirc_2013_arch}
	\caption{Architecture of the ADPLL network and of a single node \cite{zianbetov2013distributed}.}
	\label{fig:eldar_node}
\end{figure}

Looking at the design of a given node it is notable that the function carried out by the Error Combiner is akin to a average, therefore as mentioned by Pratt and Nyugen, there is potential to a mode locked equilibrium in which the oscillators are not synchronised. In their paper, they presented a method where initial start-up was performed uni-directionally and, once all nodes are close to alignment, full connectivity could be restored, however this was not viable in an analogue system as reconfigurability was not an option \cite{pratt1995distributed}. In creating an entirely digital system, Zianbetov and Shan could exploit reconfigurability and implement a uni-directional start-up and thus avoid the problem of convergence into a mode locked state, without having to design a non-linear phase detector.

\section{ADPLL Architecture}
As indicated in Figure \ref{fig:mulkeen_pll} the three main building blocks of a conventional \ac{PLL} are the \ac{PD}, \ac{LF} and \ac{VCO}. In an \ac{ADPLL} these blocks are then replaced by their digital counterparts, necessitating quantisation in order to remain physically realisable. 
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.5]{../tex_files/mulkeen_pll}
	\caption{Block Diagram of a \acl{PLL}, \textit{Wireless Systems Notes}, B. Mulkeen (2017).}
	\label{fig:mulkeen_pll}
\end{figure}

\subsection{Digitally Controllable Oscillator}
In a digital system there are a very limited number of voltages representable, most commonly just two, so using a voltage to control the oscillator is not a viable strategy. Instead a fixed bit width signal is used to control the oscillator's period, selecting the number of inverters in a ring oscillator or the varactor configuration of a travelling wave oscillator \cite{chen2011rotary}. The decisions made in the design of the \ac{DCO}, or \ac{NCO}, determine many of the other \ac{ADPLL} parameters. While tuning range and centre frequency, as well as linearity, carry over from the analogue counterpart, a \ac{DCO} also has a frequency step which in combination with the bit width of the control signal determines the range over which the oscillator can be tuned. Figure \ref{fig:my_ring} illustrates a basic ring oscillator design. A ring oscillator is an inherently unstable circuit composed of an odd number of inverters connected in a circle, which allows a signal to propagate infinitely, with the signal at any point in the circuit appearing as a square wave. The half-period of this oscillator is the time taken for the signal to propagate once through the chain, $n$ times the propagation delay through one inverter. The frequency of operation can then be set by modulating the length of the chain, in steps of two inverters to maintain an odd number, by means of \texttt{f\_select}. The main impact of output frequency quantisation is that only frequencies which are integer multiples of the frequency step away from the centre frequency can be easily reproduced, with intermediate values only obtainable in a manner akin to Fractional-N synthesis with the control code toggling back and forth. This acts as a source of jitter in the system.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.275]{../inverter_chain}
	\caption{Basic Ring/Inverter Chain Oscillator.}
	\label{fig:my_ring}
\end{figure}

It is also possible to implement a \ac{NCO} by means of a counter, in a manner that will produce either linear period or linear frequency steps. The former is achieved by setting the reload point of the counter

\subsection{Digital Phase Detector}
Once again quantisation impacts the Phase Detector, as rather than a continuous output the phase detector in an ADPLL has a finite number of output values, thus limiting the accuracy of the phase detector. A second form of quantisation is also present, as unlike an analog system, a digital phase detector does not provide continuous data in the time domain either, instead relying on sampling. At its most basic, a digital phase comparator may only output an indication of which signal is leading, a design known as a Bang-Bang Detector, which can be constructed using a single D Flip Flop. As the output only has two levels the resultant word is only 1 bit wide and as such, limits the range over which the output frequency can be controlled. More complex designs such as that in Figure \ref{fig:shan_bb_pd}, implemented by Shan, build on this by measuring the time difference between edges of the signals using a Time-to-Digital Converter, or TDC, in his case using a tapped delay line \cite{shan2014phd}. This essentially mimics a counter and allows for a non binary form of phase measurement.

\begin{figure}[h] %TODO top
	\centering
	\includegraphics[scale=0.35]{../shan_bb_pd}
	\caption{Bang-bang phase/frequency detector architecture. \cite{shan2014phd}.}
	\label{fig:shan_bb_pd}
\end{figure}
\subsection{Digital Loop Filter}
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.45]{../pi_simple} %TODO this diagram
	\caption{Basic PI Controller Architecture.}
	\label{fig:my_simple_pi}
\end{figure}
The loop filter in an ADPLL can be implemented as a PI controller as it is just a low pass filter, such as that in Figure \ref{fig:my_simple_pi}. In the case of an ADPLL node in a network the input of this filter is a weighted sum of two to four phase errors, each corresponding to a neighbouring node. In a digital system the proportional section is implemented by a simple multiplier, whereas the integral path is a combination of a multiplier and a delayed summation stage. The value of these gains determine the response and stability of the oscillator system. The transfer function of such a circuit is given by \cite{shan2014phd}:
\begin{equation*}
	H(z) = \alpha + \beta\frac{1}{1-z^{-1}} = \frac{(\alpha + \beta) - \alpha z^{-1}}{1-z^{-1}}
\end{equation*}
It has been found by Koskin \textit{et al} (2018) that stable operation can be achieved when the integral gain, $k_i$, is less than half the proportional gain, $k_p$ \cite{koskin2018generation}. In the same study a range of values was found which would produce low jitter operation of the network. As these values are all less than one, the filter must implement fixed point arithmetic in an effort to maintain the simplicity of the clock distribution network, rather than incurring the complexity penalty of floating point calculations.

\section{The Role of FPGAs}
